Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization

### 2024-09-04 manuscript.Rmd
critize Lakens definition in his recent paper

---
title: "Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization"
bibliography: references.bib
abstract: "`r paste0(c('\\\\noindent', readLines('abstract.md')), collapse = ' ')`"
shorttitle: "The Objective of Preregistration"
author: 
  - name: Aaron Peikert
    affiliation: "1,2,3"
    email: peikert@mpib-berlin.mpg.de
    corresponding : yes    # Define only one corresponding author
    address: "Center for Lifespan Psychology, Max Planck Institute for Human Development, Lentzeallee 94, 14195 Berlin, Germany"
    role:
      - Conceptualization
      - Writing---Original Draft Preparation
      - Writing---Review & Editing
      - Methodology
      - Formal analysis
      - Software
      - Visualization
      - Project administration
  - name: Maximilian S. Ernst
    affiliation: "1, 4"
    role:
      - Writing---Review & Editing
      - Formal analysis
      - Validation
  - name: Andreas M. Brandmaier
    affiliation: "1, 3, 5"
    role:
      - Writing---Review & Editing
      - Supervisions
affiliation:
  - id            : "1"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "2"
    institution   : "Department of Imaging Neuroscience, University College London, London, UK"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
keywords: |
  preregistration; confirmation; exploration; hypothesis testing; Bayesian; Open Science
wordcount         : "`r tryCatch(wordcountaddin::word_count(), error = function(x)'~8390')`"
figsintext: true
figurelist: false
tablelist : false
footnotelist: false
lineno: true
linkcolor: "blue"
csquotes: false
class: "man"
documentclass: "apa7"
csl: "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
output:
  papaja::apa6_pdf:
    latex_engine: xelatex
header-includes: |
  \raggedbottom
  \makeatletter

@@

This is the result of researchers applying one of two strategies to evade the self-imposed restrictions of preregistrations: writing a loose preregistration to begin with [@stefanBigLittleLies2022] or deviating from the preregistration afterward [@lakensWhenHowDeviate2024].
The latter is a frequent occurrence and, perhaps more worryingly, often remains undisclosed [@akkerEffectivenessPreregistrationPsychology2023; @claesenComparingDreamReality2021].
Both strategies may be used for sensible scientific reasons or with the self-serving intent of generating desirable results.
Thus, insisting on equating preregistration and confirmation has led to the criticism that, all things considered, preregistration is actually harmful and neither sufficient nor necessary for doing good science [@szollosiPreregistrationWorthwhile2020; @phamPreregistrationNeitherSufficient2021].

@@

Theoretical risk may conceptually be related to the framework of "severity" [@mayoStatisticalInferenceSevere2018; @mayoErrorStatistics2011].
Severity, is a Neopopperian view which asserts that there is evidence for a hypothesis just to the extent that it survives stringent scrutiny.
However, there are crucial differences between the two.
First, our perspective on theoretical risk is not primarily concerned with avoiding inductive reasoning but with subjective changes of belief.
This is important because, while severity is calculable, it remains unclear how severity should be valued, e.g. if an increase in severity from .80 to .81 should be as impressive as from .99 to .999.
Second, severity considerations are mainly after the fact.
Severity, a measure with which we can rule out alternative explanations, can only be calculated after evidence was observed.
This makes it difficult to guide a priori decisions in planning a study, after all severity disregards power, if we observe evidence, and disregards Type I error rate when we do not.
This implies that for a priori balancing Type I and Type II error rate, a researcher must assign a priori probabilities to, for example, the size of an effect.
Since such a move not in line with frequentist rationale there is no guidelines available on how to do this.
Third, we would argue that severity considerations assume full information about how the evidence came about and hence imply axiomatically the need for perfect preregistration.
This comes down to frequentist understanding of probability as the outcome of a well defined random experiments.
When judging a particular study, a frequentist, and hence a severe tester, may not assign probability to the event that the researchers did, for example, p-hack.
The lack of knowledge on the readers side does not turn the p hacking into a random event of which we can calculate the long run frequency aka frequentist probability.
A severe test, hence, must assume that they know the Type I and Type II error rate precisely.
Full transparency, is hence assumed, and we can not imagine many ways except preregistration that get close to this ideal.
This assumptions also makes it difficult to deal with less than perfect preregistrations and post-hoc changes without appealing to principles outside the core philosophy of severity.

However, there also are communalities, like the strong emphasis on counterfactual consideration (imagining the hypothesis was false), and there are even proposals to reconcile Bayesian and severity considerations [@vandongenBayesianPerspectiveSeverity2023].

>>>>

---
title: "Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization"
bibliography: references.bib
abstract: "`r paste0(c('\\\\noindent', readLines('abstract.md')), collapse = ' ')`"
shorttitle: "The Objective of Preregistration"
author: 
  - name: Aaron Peikert
    affiliation: "1,2,3"
    email: peikert@mpib-berlin.mpg.de
    corresponding : yes    # Define only one corresponding author
    address: "Center for Lifespan Psychology, Max Planck Institute for Human Development, Lentzeallee 94, 14195 Berlin, Germany"
    role:
      - Conceptualization
      - Writing---Original Draft Preparation
      - Writing---Review & Editing
      - Methodology
      - Formal analysis
      - Software
      - Visualization
      - Project administration
  - name: Maximilian S. Ernst
    affiliation: "1, 4"
    role:
      - Writing---Review & Editing
      - Formal analysis
      - Validation
  - name: Andreas M. Brandmaier
    affiliation: "1, 3, 5"
    role:
      - Writing---Review & Editing
      - Supervisions
affiliation:
  - id            : "1"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "2"
    institution   : "Department of Imaging Neuroscience, University College London, London, UK"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
keywords: |
  preregistration; confirmation; exploration; hypothesis testing; Bayesian; Open Science
wordcount         : "`r tryCatch(wordcountaddin::word_count(), error = function(x)'~8390')`"
figsintext: true
figurelist: false
tablelist: false
footnotelist: false
lineno: true
linkcolor: "blue"
csquotes: false
class: "man"
documentclass: "apa7"
csl: "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
output:
  papaja::apa6_pdf:
    latex_engine: xelatex
header-includes: |
  \raggedbottom
  \makeatletter

@@

This is the result of researchers applying one of two strategies to evade the self-imposed restrictions of preregistrations: writing a loose preregistration to begin with [@stefanBigLittleLies2022] or deviating from the preregistration afterward [@lakensDeviate2024].
The latter is a frequent occurrence and, perhaps more worryingly, often remains undisclosed [@akkerEffectivenessPreregistrationPsychology2023; @claesenComparingDreamReality2021].
Both strategies may be used for sensible scientific reasons or with the self-serving intent of generating desirable results.
Thus, insisting on equating preregistration and confirmation has led to the criticism that, all things considered, preregistration is actually harmful and neither sufficient nor necessary for doing good science [@szollosiPreregistrationWorthwhile2020; @phamPreregistrationNeitherSufficient2021].

@@

Theoretical risk may conceptually be related to the framework of "severity" [@mayoStatisticalInferenceSevere2018; @mayoErrorStatistics2011].
Severity, is a Neopopperian view which asserts that there is evidence for a hypothesis just to the extent that it survives stringent scrutiny.
However, there are crucial differences between the two.
First, our perspective on theoretical risk is not primarily concerned with avoiding inductive reasoning but with subjective changes of belief.
This is important because, while severity is calculable, it remains unclear how severity should be valued, e.g. if an increase in severity from .80 to .81 should be as impressive as from .99 to .999.
Second, severity considerations are mainly after the fact.
Severity, a measure with which we can rule out alternative explanations, can only be calculated after evidence was observed.
This makes it difficult to guide a priori decisions in planning a study, after all severity disregards power, if we observe evidence, and disregards Type I error rate when we do not.
This implies that for a priori balancing Type I and Type II error rate, a researcher must assign a priori probabilities to, for example, the size of an effect.
Since such a move is not in line with frequentist rationale there is no guidelines available on how to do this.
Third, we would argue that severity considerations assume full information about how the evidence came about and hence imply axiomatically the need for perfect preregistration.
This comes down to frequentist understanding of probability as the outcome of a well defined random experiments.
When judging a particular study, a frequentist, and hence a severe tester, may not assign probability to the event that the researchers did, for example, p-hack.
The lack of knowledge on the readers side does not turn the p hacking into a random event of which we can calculate the long run frequency aka frequentist probability.
A severe test, hence, must assume that they know the Type I and Type II error rate precisely.
Full transparency, is hence assumed, and we can not imagine many ways except preregistration that get close to this ideal.
This assumptions also makes it difficult to deal with less than perfect preregistrations and post-hoc changes without appealing to principles outside the core philosophy of severity.
One such approach is @lakensDeviate2024' introduction of validity as an additional consideration to severity when evaluating deviations from preregistrations.
Interestingly, in this work he unconventionally defines high severity as high $P(E|H)$ and high $P(\neg{}E|\neg{}H)$, which is closer to definitions of "diagnosticity" [@fiedlerWhatConstitutesStrong2017; @oberauerAddressingTheoryCrisis2019] and falls under the broad class of measures for evaluating evidence we consider here.
Notably, the original definition of severity does not satisfy the statistical relevancy condition and is not such a measure; @mayoStatisticalInferenceSevere2018, p. 14:

> Severity Principle (strong): We have evidence for a claim C just to the extent it survives a stringent scrutiny. If C passes a test that was highly capable of findings flaws or discrepancies from C, and yet none or few are found, the passing result, x, is evidence for C.

However, there also are communalities between our approach and severity, like the strong emphasis on counterfactual consideration (imagining the hypothesis was false), and there are even proposals to reconcile Bayesian and severity considerations [@vandongenBayesianPerspectiveSeverity2023].
